---
title: "The Datafather: Essential Coverage & Life Expectancy Saga"
subtitle: "Exploring the Evolutionary Dynamics Across Nations: How Economic Prosperity, Healthcare Accessibility, and Longevity Intertwine Across Time"
author: "Antonio Gogioman"
date: "04-04-2024"
format:
  html:
    theme: journal
    backgroundcolor: lightgrey
    toc: true
execute: 
  echo: false
  warning: false
  message: false
---
## Introduction

In recent centuries, the dynamics of **health coverage**, **GDP Per Capita** *(Gross Domestic Product)*, and **life expectancy** have undergone profound changes. Leveraging ***UNICEF data***, we explore these transformations. Historically, limited healthcare access gave way to broader coverage, while economic growth increasingly correlated with improved health outcomes. Concurrently, life expectancy surged, reflecting advancements in healthcare and socioeconomic conditions. Through ***concise analysis***, we aim to uncover key insights into these interconnected trends, guiding future health and economic policies toward greater global well-being.

To display the relevant information, graphs will indicate key insights and display correlations between the figures. The graphs utillised are as follows:

1. Map Graph
2. Time Series Graph
3. Scatter Plot Graph
4. Bar Chart

GDP Per Capita
: The average economic output per person in a country, indicating individual economic prosperity within the nation.

Health Coverage
: The accessibility and affordability of healthcare services for a nation's population, reflecting its commitment to ensuring comprehensive medical care for all citizens.

## Report

```{r}
#| label: setup
#| include: false

# libraries
library(tidyverse)
library(plotly)
library(ggplot2)

# data
unicef_metadata <- read_csv("unicef_metadata.csv")
unicef_indicator_2 <- read_csv("unicef_indicator_2.csv")

#transformations
data_join <- full_join(unicef_metadata, unicef_indicator_2, by = c("country" = "country", "year" = "time_period"))
country_agg <- aggregate(GDPPerCap ~ country, data_join, mean)

country_agg <- country_agg[order(-country_agg$GDPPerCap), ]
```

### World Map Graph
```{r}
#| fig-cap: "This world map illustrates life expectancy across different countries in 2005. Each country is color-coded according to its life expectancy, with darker blues representing longer life expectancies and lighter red shades indicating shorter ones. This visualization offers insights into global disparities in life expectancy and highlights regional variations in health outcomes."


map_world <- map_data("world")

data_join_2005 <- data_join %>%
  filter(year == 2005)

map_data_join_2005 <- full_join(data_join_2005, map_world, by = c("country" = "region"))

ggplot(map_data_join_2005) +
  aes(x = long, y = lat, group = group, fill = LifeExp) +
  geom_polygon(color = "black") + # Add country borders +
  scale_fill_gradient(low = "red", high = "blue", name = "Life Expectancy") +
  labs(title = "Life Expectancy by Country in 2005",
       caption = "Life Expectancy Graph",
       fill = "Life Expectancy") +
  theme_minimal() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        plot.title = element_text(size = 14, face = "bold"),
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom")

```

Transitioning from the world map graph to the time series, we shift our focus from a snapshot of life expectancy disparities among countries to a dynamic exploration of how these trends have evolved over time. As we navigate from the macro-level insights of the map to the micro-level insights of the time series, we gain a deeper understanding of the **intricate factors influencing** global health outcomes throughout history.

### Time Series Graph
```{r}
#| fig-cap: "This interactive time series graph presents the life expectancy trends of multiple countries from 1960 onwards. The graph allows users to scroll across the ledger to view different countries' movements over time. Additionally, by double-clicking on a specific country, users can isolate and focus on that country's life expectancy trajectory. This visualization offers an engaging exploration of global health dynamics and highlights the diverse experiences of nations in improving life expectancy."
timeseries_plot_1 <- data_join %>%
  ggplot() +
  aes(year, LifeExp, colour = country) +
  geom_line() +
  labs(
    x = "Year",
    y = "Life Expectancy",
    title = "Time Series Displaying Life Expectancy by Country",
    colour = "Country"
  )

ggplotly(timeseries_plot_1)
```

The time series graph provides us with plenty of interesting insights,

* We see sudden drops in avg. life expectancy drop to around **12 years** in Cambodia in **1975**
* Similar drops are noted in South Sudan and Rwanda
* ***Hong Kong*** appears to be the country with the ***highest average life*** expectancy as of *2021*



### Scatter Plot Graph
```{r}
#| fig-cap: "This concise visualization illustrates the correlation between life expectancy and GDP per capita across countries in 2005. By comparing these two key indicators on a global scale, insights into the potential relationship between economic prosperity and health outcomes are revealed."
#| fig-cap-location: margin


data_join_2005 <- data_join %>%
  filter(year == 2005)

p <- ggplot(data_join_2005) +
  aes(GDPPerCap, LifeExp, colour = country, size = Population, text = country) +
  geom_point(alpha = 0.6) +
  facet_wrap(~ year) +
  labs(
    x = "GDP Per Capita",
    y = "Life Expectancy",
    title = "The Relationship of Life Expectancy and GDP Per Capita in 2005"
  ) +
  guides(colour = "none", size = "none") +
  theme_classic() +
  theme(text = element_text(family = "serif"))
  
p <- ggplotly(p, tooltip = "text")

p
```


Moving from the scatter plot showing average GDP per Capita and life expectancy, we transition to a temporal exploration. Now, we shift focus to a bar chart illustrating the average GDP over time. This change in visualisation offers a dynamic perspective, allowing us to track the fluctuations in economic growth throughout the years.

### Bar Chart
```{r}
#| fig-cap: "This interactive bar chart visualizes the average GDP per capita for each country over the entire time period. Upon hovering over a bar, users can zoom in to reveal the corresponding country without the need for explicit labels. This dynamic visualization offers a streamlined way to explore and compare GDP per capita trends across nations."

country_agg <- aggregate(GDPPerCap ~ country, data_join, mean)

country_agg <- country_agg[order(-country_agg$GDPPerCap), ]

horizontal_bar_plot <- ggplot(country_agg, aes(x = GDPPerCap, y = reorder(country, GDPPerCap), text = country)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    x = "Average GDP Per Capita",
    y = "Country",
    title = "Average GDP Per Capita by Country"
  ) +
  theme_classic() +
  theme(
    text = element_text(family = "serif"),
    axis.text.x = element_blank()
  ) +
  coord_flip()

horizontal_bar_plot <- ggplotly(horizontal_bar_plot, tooltip = "text")

horizontal_bar_plot
```

## Conclusion

In conclusion, the analysis highlights the **critical relationship** between *health coverage*, *GDP per capita*, and *life expectancy worldwide*. The visualizations, such as the world map and dynamic bar chart, offer succinct insights into global trends and disparities. Moving forward, targeted efforts to improve health coverage and economic opportunities are ***imperative*** for fostering inclusive development and better health outcomes globally.
